The unphysicality of Hilbert spaces

### 2023-09-03 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Updated version based on arxiv version comments

\begin{abstract}
We show that Hilbert spaces should not be considered the ``correct'' spaces to represent quantum states mathematically. We first prove that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice-versa, and time evolutions that transform finite expectations to infinite ones in finite time. This makes Hilbert spaces physically unsound as they model a potential infinity as an actual infinity. We suspect that at least some problems in quantum theory related to infinities may be ultimately caused by the wrong space being used. We strongly believe a better solution can be found, and we look at Schwartz spaces for inspiration, as, among other things, they guarantee that the expectation of all polynomials of position and momentum are finite, guarantee solution to the moment problem and are the only space closed under Fourier transform.
\end{abstract}

@@

Let us first clarify the standard we believe should be used to judge the physicality of a formal definition.\cite{redei_tension_2020, north_physics_2021} Suppose we have a physical system under study in a given condition. A mathematical definition is \textbf{physical} if it properly characterizes a physical system. That is, if it can be shown that the mathematical definition is needed to capture and only capture a particular aspect of the physical system. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart.

@@

The argument can be summed up as follows. On physical grounds, we assume the existence of quantities, like position, that can take arbitrarily large values. This requires that, in principle, for each possible value, there must exist some state whose expectation matches that value. Completeness forces us to include states with an infinite expectation value, which cannot be implemented physically. Thus the mathematical definition implicitly assumes objects that cannot be physically realized, so \textbf{the use of completeness for the space of quantum systems is unphysical.}

@@

\begin{itemize}
\item its elements are fully identified by the expectations - two Schwartz functions are the same if and only if all expectations are the same; quantum tomography on all polynomials fully reconstructs the state
\item it is an inner product space with the same norm as $L^2$
\item it is a dense subset of $L^2$ - any element of the Hilbert space can be approximated by a Schwartz function with an arbitrary level of precision
\item it is complete with respect to the expectation values - a sequence of functions for which all expectation values of all polynomials of position and momentum converge will converge in the Schwartz space
\item its closure over all Cauchy sequences recovers the Hilbert space - the only thing missing are the Cauchy sequences for which the limit of at least one expectation is not well defined
\item it is the only space closed under Fourier transform - the Fourier transform of a Schwartz function is a Schwartz function
\item it plays a fundamental role in the theory of distributions - objects like the $\delta$-function are mathematically defined on top of the Schwartz space
\item the Hermite polynomials, the solutions of the harmonic oscillator, are Schwartz functions
\end{itemize}
These properties make the Schwartz space a much more reasonable candidate to capture the physics. The convergence of expectation values of actual observables replaces convergence over vectors in an abstract space.\footnote{Compare this statement with the ``weak convergence'' of von Neumann algebras.} To again be clear, we are not proposing this as a solution. We are merely pointing out that alternatives do exist and should be fully explored.\footnote{Note that $C^*$-algebras model observables and therefore do not solve the problem at hand of modeling states.}

@@

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of functions with compact support on phase space gives us finite expectation value for all functions of positions and momentum.}

@@

Not only do the expectations of all polynomials exist, but they are enough to identify the elements of the space and define limits. The Schwartz space is, in fact, a Fréchet space \cite[Theorem V.9]{reed_methods_1980}, meaning that it is a Hausdorff space, with the topology induced by the seminorms given by the expectations, and it is complete with respect to those seminorms. Physically, this means that measurements on the polynomials (i.e. the topology of the seminorms) are enough to identify the states (i.e. the singletons, which are closed sets as per Hausdorff).\footnote{This is not true in classical mechanics. Even in the single variable case, distinct distributions can have the same moments. However, if we assume compact support, distinct distributions have distinct moments even in the multi-variable case.\cite{moment_problem_2017} It is curious that the moment problem in quantum mechanics works differently.} Moreover, if we have a sequence of states for which all expectations converge, we are guaranteed to converge to another state (i.e. completeness with respect to the seminorms). If any expectation diverges, the limit is not in the space. This criterion of convergence seems a lot more physically sound.

@@

Though we believe the arguments to be thoroughly convincing, we are stopping short of declaring Schwartz spaces physical. First, this is not the goal of the paper, which is to show the unphysicality of Hilbert spaces. Second, we would need to show the importance of expectation values of polynomials from experimental considerations.\footnote{The fact that the moment problems work differently in classical and quantum mechanics may make that argument difficult to construct.} It is not clear to us that that's the case, so it may be that the justification would have to proceed in a different, but ultimately equivalent, way.\footnote{One of the issues is the fact that uniqueness of the moment problem works differently for classical and quantum mechanics.}

>>>>

\begin{abstract}
We show that Hilbert spaces should not be considered the ``correct'' spaces to represent quantum states mathematically. We first prove that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice-versa, and time evolutions that transform finite expectations to infinite ones in finite time. This makes Hilbert spaces physically unsound as they model a potential infinity as an actual infinity. We suspect that at least some problems in quantum theory related to infinities may be ultimately caused by the wrong space being used. We strongly believe a better solution can be found, and we look at Schwartz spaces for inspiration, as, among other things, they guarantee that the expectation of all polynomials of position and momentum are finite, their elements are uniquely identified by these expectations, and they are the only space closed under Fourier transform.
\end{abstract}

@@

Let us first clarify the standard we believe should be used to judge the physicality of a formal definition.\cite{redei_tension_2020, north_physics_2021} Suppose we have a physical system under study in a given condition. A mathematical definition is \textbf{physical} if it properly characterizes the physical system. That is, if it can be shown that the mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart.

@@

The argument can be summed up as follows. On physical grounds, we assume the existence of quantities, like position, that can take arbitrarily large values. This requires that, in principle, for each possible value, there must exist some state whose expectation matches that value. Completeness forces us to include states with an infinite expectation value, which cannot be implemented physically. Thus the mathematical definition implicitly assumes objects that cannot be physically realized, so \textbf{the use of completeness for the space of quantum systems is unphysical.}\footnote{As rigged Hilbert spaces are Hilbert spaces with additional structure, they cannot solve this problem. For further discussion, see the appendix.}

@@

\begin{itemize}
\item its elements are fully identified by the expectations - two Schwartz functions are the same if and only if all expectations are the same; quantum tomography on all polynomials fully reconstructs the state
\item it is an inner product space with the same norm as $L^2$
\item it is a dense subset of $L^2$ - any element of the Hilbert space can be approximated by a Schwartz function with an arbitrary level of precision
\item it is complete with respect to the expectation values - a sequence of functions for which all expectation values of all polynomials of position and momentum converge will converge in the Schwartz space
\item its closure over all Cauchy sequences recovers the Hilbert space - the only thing missing are the Cauchy sequences for which the limit of at least one expectation is not well defined
\item it is the only space closed under Fourier transform - the Fourier transform of a Schwartz function is a Schwartz function
\item it plays a fundamental role in the theory of distributions - objects like the $\delta$-function are mathematically defined on top of the Schwartz space
\item the Hermite polynomials, the solutions of the harmonic oscillator, are Schwartz functions
\end{itemize}
These properties make the Schwartz space a much more reasonable candidate to capture the physics.\cite{albert2022bosonic} The convergence of expectation values of actual observables replaces convergence over vectors in an abstract space.\footnote{Compare this statement with the ``weak convergence'' of von Neumann algebras.} To again be clear, we are not proposing this as a solution. We are merely pointing out that alternatives do exist and should be fully explored.\footnote{Note that $C^*$-algebras model observables and therefore do not solve the problem at hand of modeling states.}

@@

\section{The unphysicality of rigged Hilbert spaces}

As part of the feedback we received on previous drafts, some people argued that rigged Hilbert spaces were developed to solve exactly the problems we highlight. They were not. They are designed to solve another problem, namely extending the spectral theorem to the case of continuous spectra.\cite{Madrid_2005}

First of all, note that ``rigged'' in this context does not mean ``manipulated by dishonest intention'', but rather ``fully equipped and ready to go'' as, for example, a fully-rigged ship. Therefore, the name already tells us that rigged Hilbert spaces are Hilbert spaces with additional structure. They still satisfy completeness, and thus are affected by the problems presented above. The unphysicality of a mathematical structure cannot simply be solved by adding more structure: the unphysical elements must be removed.

Rigged Hilbert spaces exist to address the fact that an operator may have eigenfunctions that are not in the Hilbert space. For example, the eigenstates for the operator $P = - \imath \hbar \partial_x$ are of the form $e^{\frac{xp}{-\imath\hbar}}$, and are not square integrable. This is due to the continuous spectrum, not the unboundedness of the operator, and is therefore an independent problem. We can show that with two examples.

Bounded operators can have a continuous spectrum. For example, let $\mathcal{H}$ be the Hilbert space of all wave functions defined on the compact domain $x \in [0,1]$. The position operator, in this case, is bounded. We do not have problems of infinite expectation value. However, the position operator $X$ does not have eigenfunctions in $\mathcal{H}$. The delta Dirac distribution, in fact, is not part of the Hilbert space; a sequence of Gaussians with decreasing variance, whose limit is a delta Dirac, is not a Cauchy sequence. Therefore we cannot properly talk about eigenstates of $X$ in $\mathcal{H}$.

Unbounded operators can have a discrete spectrum. For example, the harmonic oscillator has a discrete spectrum, given by the Hermite polynomials. All eigenfunctions are in the Hilbert space.

Let us give a very rough description of the rigged Hilbert space construction. Given a generic Hilbert space $\mathcal{H}$, we take an operator $X$, which we assume unbounded. The domain $D(X)$ of the operator is a strict subset of $\mathcal{H}$. That is, $D(X) \subset \mathcal{H}$. We can then define $D(X)^*$ as the space of linear functionals of $D(X)$.\footnote{A similar construction can be done with anti-linear functionals, giving the anti-dual space.} Mathematically, $D(X)$ is the space of test functions while $D(X)^*$ is the space of tempered distributions, the dual of $D(X)$. This is a strict superset of $\mathcal{H}$. That is,
\begin{equation}
	D(X) \subset \mathcal{H} \subset D(X)^*.
\end{equation}
This is called the Gelfand triple. This construction allows us to talk about $\delta(x)$, as it is a tempered distribution, and therefore it can be found in $D(X)^*$. Note that if we chose a different operator, say $P$, we would have a different construction and a different rigged Hilbert space.

The point of showing the construction is to see that the rigged Hilbert space is a Hilbert space plus two other spaces. Therefore, a rigged Hilbert space is physical only if all three spaces are physical. If the Hilbert space is not physical, neither is the rigged Hilbert space.

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true for distributions in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of distributions with compact support on phase space gives us finite expectation value for all functions of positions and momentum.}

@@

Not only do the expectations of all polynomials exist, but they are enough to identify the elements of the space and define limits. The Schwartz space is, in fact, a Fréchet space \cite[Theorem V.9]{reed_methods_1980}, meaning that it is a Hausdorff space, with the topology induced by the seminorms given by the expectations, and it is complete with respect to those seminorms. Physically, this means that measurements on the polynomials (i.e. the topology of the seminorms) are enough to identify the states (i.e. the singletons, which are closed sets as per Hausdorff).\footnote{This essentially tells us that the moment problem restricted to pure states has a unique solution in quantum mechanics. Setting up a comparison with the classical case is delicate. If we restrict ourselves to classical pure states, i.e. points in phase space, then the solution is also unique. If we restrict ourselves to compact support, distinct distributions have distinct moments even in the multi-variable case.\cite{moment_problem_2017} If we restrict ourselves to integrable functions, even in the single variable case, distinct distributions can have the same moments. None of these comparisons seems adequate as a quantum state is a statistical object by nature, but it is not an arbitrary mixture. A more proper comparison would seem to restrict the problem to the space of integrable functions with zero entropy.} Moreover, if we have a sequence of states for which all expectations converge, we are guaranteed to converge to another state (i.e. completeness with respect to the seminorms). If any expectation diverges, the limit is not in the space. This criterion of convergence seems a lot more physically sound.

@@

Though we believe the arguments to be thoroughly convincing, we are stopping short of declaring Schwartz spaces physical. First, this is not the goal of the paper, which is to show the unphysicality of Hilbert spaces. Second, we would need to show the importance of expectation values of polynomials from experimental considerations.\footnote{The lack of a clear classical parallel for the moment problem restricted to pure states may make that argument difficult to construct.} It is not clear to us that that's the case, so it may be that the justification would have to proceed in a different, but ultimately equivalent, way.
